---
title: "Introduction to Elmasri et al.(2020) Host-Parasite Link Prediction Package"
author: "Maxwell J. Farrell & Mohamad Elmasri"
date: "`r format(Sys.time(), '%B %d %Y')`"
output:
    rmarkdown::html_vignette:
        toc: yes
        toc_depth: 3
        number_sections: yes
    pdf_document:
        toc: true
        toc_depth: 3
        number_sections: yes
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction
```{r setup, include = FALSE, eval=TRUE, echo=TRUE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```
```{r loadinglibrary}
library(HPprediction)
```

## Data

To begin we will use a subsection of the [GMPD 2.0](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecy.1799) and the associated Mammal supertree updated by [Fritz et al. 2009](https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1461-0248.2009.01307.x)

```{r loadData, eval=TRUE, echo = TRUE}
data(gmpd)
data(mammal_supertree)
head(gmpd)
print(mammal_supertree)

# Removing parasites not reported to species
gmpd <- gmpd[grep("sp[.]",gmpd$ParasiteCorrectedName, invert=TRUE),]
gmpd <- gmpd[grep("not identified",gmpd$ParasiteCorrectedName, invert=TRUE),]

# Subsetting host family Bovidae
gmpd <- gmpd[gmpd$HostFamily=="Bovidae",]

# Formatting host names to match phylogeny
gmpd$HostCorrectedName <- gsub(" ","_", gmpd$HostCorrectedName)

# Creating binary interaction matrix
com <- table(gmpd$HostCorrectedName, gmpd$ParasiteCorrectedName)
com[com>1] <- 1
com <- as.matrix(unclass(com), nrow=nrow(com), ncol=ncol(com))

# loading phylogeny and pruning to hosts in the interaction matrix
mammal_supertree <- drop.tip(mammal_supertree, mammal_supertree$tip.label[!mammal_supertree$tip.label%in%rownames(com)])

# merge the tree and interaction matrix
cleaned <- network_clean(com, mammal_supertree, 'full')
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree
```


## Plotting Input Data

To visualize the structure of the input interaction matrix we can use two built-in plotting functions.

```{r plotZ, echo=TRUE, eval=TRUE,fig.align='center'}
plot_Z(com, tickMarks=20, cex.lab=1, cex.axis=1)

```

We can also use the function `lof` to left order the interaction matrix before plotting:

```{r plotZ_lof, eval=TRUE, echo=TRUE,fig.align='center'}
plot_Z(lof(com), tickMarks=20, cex.lab=1, cex.axis=1)

```

Another interesting property of networks to explore is the degree distribution. We can look at the degree distribution for both hosts and parasites with the following function: 

```{r plotDD, eval=TRUE, echo=TRUE,fig.align='center'}
plot_degree(com, cex.lab=1, cex.axis=1, pt.cex=1, legend.cex=1)

```


## Running the model

Running for 1500 slices (iterations) on an i7-6560U and 16GB RAM took about 25 seconds. 

```{r network_est, eval=TRUE, echo=TRUE}
out <- network_est(Z = com, tree = tree, slices = 1000, model.type = 'full')
str(out)
```

### Example traceplots

```{r traceplots, echo=TRUE, eval=TRUE}
# Affinity parameter of parasite 1
plot(out$param$w[1,],type="l", col=2, ylab="Affinity parameter for parasite 1", xlab="Iteration")

# Affinity parameter of host 1
plot(out$param$y[1,],type="l", col=4, ylab="Affinity parameter for host 1", xlab="Iteration")

# Phylogeny scaling parameter
plot(out$param$eta, type="l",col="darkgreen", ylab="Phylogeny scaling parameter (EB model)", xlab="Iteration")
```


### Generating posterior probability matrix

```{r sample_param, eval=TRUE, echo=TRUE}
P <- sample_parameter(param=out$param, MODEL="full", Z=out$Z, tree=out$tree, size = 500)
```

### Identfying top predicted links with no documentation

```{r topundoc, echo=TRUE, eval=TRUE}
topPairs(P, out$Z)
```

## General approach to running the model with 5-fold cross validation

```{r cv, echo=TRUE, eval=FALSE}
## General variables
MODEL = 'full'                       # full, distance or affinity
SLICE = 1000                          # no of iterations
NO.CORES = 3                         # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3					 # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3					 # hyperparameter for prior over columns affinity, effective under affinity and full models only

## Loading required packages
require(parallel)

## preparing tree and com
cleaned <- network_clean(com, tree, 'full')
com <- cleaned$Z                         # cleaned binary interaction matrix
tree <- cleaned$tree                     # cleaned tree

## indexing 5-folds of interactions
folds <- cross.validate.fold(com, n= 5, min.per.col=2)  
[1] "Actual cross-validation rate is 0.095"
[2] "Actual cross-validation rate is 0.095"
[3] "Actual cross-validation rate is 0.095"
[4] "Actual cross-validation rate is 0.095"
[5] "Actual cross-validation rate is 0.096"

# returns a matrix of 3 columns (row, col, group), (row, col) correspond to Z, group to the CV group
tot.gr <- length(unique(folds[,'gr']))   # total number of CV groups

## A loop to run over all CV groups
res <- mclapply(1:tot.gr, function(x, folds, Z, tree, slice, model.type, ALPHA.ROWS, ALPHA.COLS){

    ## Analysis for a single fold
    Z.train = Z
    Z.train[folds[which(folds[,'gr']==x),c('row', 'col')]]<-0

    ## running the model of interest
    obj = network_est(Z.train, slices=slice, tree=tree, model.type=model.type,
                      a_y = ALPHA.ROWS, a_w = ALPHA.COLS)

    P = sample_parameter(obj$param, model.type, Z.train, tree)
    Eta = if(is.null(obj$param$eta)) 0 else mean(obj$param$eta)
    
    ## order the rows in Z.test as in Z.train
    roc = rocCurves(Z, Z.train, P, plot=FALSE, bins=400, all=FALSE)
    tb  = ana.table(Z, Z.train, P, roc,  plot=FALSE)
    roc.all = rocCurves(Z, Z.train, P=P, plot=FALSE, bins=400, all=TRUE)
    tb.all  = ana.table(Z, Z.train, P, roc.all, plot=FALSE)
    
    list(param=list(P=P, Eta = Eta), tb = tb,
         tb.all = tb.all, FPR.all = roc.all$roc$FPR,
         TPR.all=roc.all$roc$TPR, FPR = roc$roc$FPR, TPR=roc$roc$TPR)
    
},  
	folds=folds, Z = com, tree=tree, model.type=MODEL, slice = SLICE,
    ALPHA.ROWS = ALPHA.ROWS, ALPHA.COLS= ALPHA.COLS, 
    mc.preschedule = TRUE, mc.cores = min(tot.gr, NO.CORES))
[1] "Running full model..."
[1] "Running full model..."
[1][1] "Running full model..."
 "Run for 1000 slices with 500 burn-ins"
[1] "Matrix dimension: 70 x 442"
[1] "Run for 1000 slices with 500 burn-ins"
[1] "Matrix dimension: 70 x 442"
[1] "Run for 1000 slices with 500 burn-ins"
[1] "Matrix dimension: 70 x 442"
[1] "slice: 200, at 2020-04-18 18:08:37"
[1] "slice: 200, at 2020-04-18 18:08:37"
[1] "slice: 200, at 2020-04-18 18:08:37"
[1] "slice: 400, at 2020-04-18 18:08:41"
[1] "slice: 400, at 2020-04-18 18:08:41"
[1] "slice: 400, at 2020-04-18 18:08:41"
[1] "slice: 600, at 2020-04-18 18:08:45"
[1] "slice: 600, at 2020-04-18 18:08:45"
[1] "slice: 600, at 2020-04-18 18:08:45"
[1] "slice: 800, at 2020-04-18 18:08:51"
[1] "slice: 800, at 2020-04-18 18:08:51"
[1] "slice: 800, at 2020-04-18 18:08:51"
[1] "slice: 1000, at 2020-04-18 18:08:57"
[1] "Done!"
[1] "slice: 1000, at 2020-04-18 18:08:57"
[1] "Done!"
[1] "slice: 1000, at 2020-04-18 18:08:58"
[1] "Done!"
[1] "Running full model..."
[1] "Run for 1000 slices with 500 burn-ins"
[1] "Matrix dimension: 70 x 442"
[1] "Running full model..."
[1] "Run for 1000 slices with 500 burn-ins"
[1] "Matrix dimension: 70 x 442"
[1] "slice: 200, at 2020-04-18 18:09:07"
[1] "slice: 200, at 2020-04-18 18:09:07"
[1] "slice: 400, at 2020-04-18 18:09:12"
[1] "slice: 400, at 2020-04-18 18:09:12"
[1] "slice: 600, at 2020-04-18 18:09:18"
[1] "slice: 600, at 2020-04-18 18:09:19"
[1] "slice: 800, at 2020-04-18 18:09:25"
[1] "slice: 800, at 2020-04-18 18:09:26"
[1] "slice: 1000, at 2020-04-18 18:09:32"
[1] "Done!"
[1] "slice: 1000, at 2020-04-18 18:09:32"
[1] "Done!"
```

We can analyze the performance of the model via the area under the receiver operating characteristic curve (AUC), and the proportion of 1s in the original data successfully recovered. 

```{r echo=FALSE, eval = TRUE}
res = readRDS('rescv.rds')
MODEL = 'full'                       # full, distance or affinity
SLICE = 1000                          # no of iterations
NO.CORES = 3                         # maximum cores to use
COUNT = TRUE                         # TRUE = count data, FALSE = year of first pub.
ALPHA.ROWS = 0.3					 # hyperparameter for prior over rows affinity, effective under affinity and full models only
ALPHA.COLS = 0.3					 # hyperparameter for prior over columns affinity, effective under affinity and full models only

## Loading required packages
require(parallel)
```

```{r diagnostics, eval=TRUE, echo=FALSE}
## Some analysis results, AUC, %1 recovered
TB = data.frame(
    m.auc = sapply(res, function(r) r$tb$auc),
    m.pred.held.out.ones = sapply(res,function(r) r$tb$pred.held.out.ones),
    m.thresh = sapply(res, function(r) r$tb$thresh),
    m.hold.out = sapply(res, function(r) r$tb$held.out.ones)
)
TB

## Printing and writing out average MCMC 
print(sprintf('Model: %s, AUC: %f and percent 1 recovered from held out: %f',
              MODEL,mean(TB$m.auc), mean(TB$m.pred.held.out.ones)))

## ROC curve points, can plot as plot(ROCgraph)
ROCgraph = cbind(
    FPR = rowMeans(sapply(res, function(r) r$FPR)),
    TPR = rowMeans(sapply(res, function(r) r$TPR)))

plot(ROCgraph, type="l", lty=2, col=2)

```

We can also construct the posterior probability matrix 'P' as the average across each fold, and look at the top undocumented interactions.

```{r topundoc_cv, eval=TRUE, echo=TRUE}
## Constructing the P probability matrix from CV results
P = matrix(rowMeans(sapply(res, function(r) r$param$P)),
    nrow = nrow(com), ncol = ncol(com))

## left ordering of interaction and probability matrix
indices = lof(com, indices = TRUE)
com = com[, indices]
P = P[, indices]
rownames(P)<-rownames(com)
colnames(P)<-colnames(com)

## view top undocumented interactions
topPairs(P,1*(com>0),topX=10)
```

We can also compare the input matrix to the posterior interaction matrix, and the orginal phylogeny compared to the phylogeny with estimated EB scaling.

```{r comparison_plots, echo=TRUE, eval=TRUE}
par(mfrow=c(1,2))

## printing input Z
plot_Z(com, tickMarks=20)

## printing posterior interaction matrix
plot_Z(1*(P > mean(sapply(res, function(r) r$tb$thres))), tickMarks=20)

## printing input tree
plot(tree, show.tip.label=FALSE)

## printing output tree
if(grepl('(full|dist)', MODEL)){
    Eta = mean(sapply(res, function(r) r$param$Eta))
    print(paste('Eta is', Eta))
    plot(rescale(tree, 'EB', Eta), show.tip.label=FALSE)
}
```

## References

Elmasri, M., Farrell, M. J., Davies, T. J., & Stephens, D. A. (2020). A hierarchical Bayesian model for predicting ecological interactions using scaled evolutionary relationships. [Annals of Applied Statistics, 14(1), 221-240.](https://projecteuclid.org/euclid.aoas/1587002672)


